---
title: "DataWrangling2EricPeter"
format: pdf
editor: visual
---

## #Data Wrangling 2 by Peter and Eric

```{r}
#Reading in our library
library(tidyverse)
library(here)
```

##Working with whale dataset

```{r}
# Read in the data
whale_popn_wide <- read_csv(here("data/whale_popn_wide.csv"))

# Take a look at it
whale_popn_wide

## Weird `` column examples:
# Ex. Fetch the data within one column
whale_popn_wide$`Gray whale`

# Ex. Plot Gray whale abundance vs Blue whale abundance
whale_popn_wide %>% 
  ggplot(aes(x = `Gray whale`,
             y = `Blue whale`)) +
  geom_point()

#Separating columns
whale_popn_wide.sep <- whale_popn_wide %>% 
  # Separate the site_year column
  separate(col = site_year,
           # Separate into two columns in this case
           into = c("site", "year"),
           # Separate by the "_"
           sep = "_"
  )

# Now look at the data
whale_popn_wide.sep

#Making year a numeric column 
whale_popn_wide.sep <- whale_popn_wide %>% 
  # Separate the site_year column
  separate(col = site_year,
           # Separate into two columns in this case
           into = c("site", "year"),
           # Separate by the "_"
           sep = "_"
  ) %>% 
  # Run the year column through the as.numeric() function and re-store that as year
  mutate(year = as.numeric(year))

# Now look at the data
whale_popn_wide.sep

#Another method for separating columns
whale_popn_wide %>% 
  # Separate the site_year column
  separate(col = site_year,
           # Separate into two columns in this case
           into = c("site", "year"),
           # Separate by the "_"
           sep = 4
  )

#Practicing unite function

whale_popn_wide.sep %>% 
  # Unite the site and year columns back into site_year
  unite(col = site_year, # New col name is going to be site_year again
        # Add in the existing columns that we want to unite together
        site, year,
        # Separate by a comma and space this time ", "
        sep = ", "
  )

#Using pivot to turn into long format function

# Pivot to long format
whale_long <- whale_popn_wide.sep %>% 
  # Take the columns we want to pivot
  pivot_longer(cols = c(`Gray whale`, `Humpback whale`, `Blue whale`),
               # Put the column names into a column called "species"
               names_to = "species",
               # Put the cell values into a column called "population_size"
               values_to = "population_size")
# Check out the data
whale_long

#Pivoting using the column number

whale_popn_wide.sep %>% 
  pivot_longer(cols = c(4:6), # c(4:6) creates a vector of the numbers 4 through 6, aka 4, 5, 6
               names_to = "species",
               values_to = "population_size")

#Choosing columns not to pivot

# Pivot to long format
whale_popn_wide.sep %>% 
  # Specify the columns we DON'T want to pivot
  pivot_longer(cols = -c(site, year, annual_temp),
               # Put the column names into a column called "species"
               names_to = "species",
               # Put the cell values into a column called "population_size"
               values_to = "population_size")

# Or:

whale_popn_wide.sep %>% 
  # Specify we DON'T want to pivot columns 1 through 3
  pivot_longer(cols = -c(1:3), # c(1:3) creates a vector of the numbers 1 through 3, aka 1,2,3
               names_to = "species",
               values_to = "population_size")

#Now let's graph this data 

whale_long %>% 
  ggplot(aes(x = year, y = population_size, color = species)) +
  geom_point() +
  geom_line()

```

##1.4 Pivot to wide format

```{r}
#Practicing pivot_wider function 

whale_long %>%
  pivot_wider(names_from = species,
              values_from = population_size)
```

##1.5 Using group_by and summarize

```{r}
whale_long %>% arrange(species, year)

#Grouping and summarizing data by species 
whale_long %>% 
  group_by(species)

#Sumarizing our grouped data

whale_long %>% 
  group_by(species) %>% 
  summarize(mean_pop = mean(population_size))

#Removing NA's in the mean calculation

whale_long %>% 
  group_by(species) %>% 
  summarize(mean_pop = mean(population_size, na.rm = TRUE))

#Calculating standard deviation

whale_long %>% 
  group_by(species) %>% 
  summarize(mean_pop = mean(population_size, na.rm = TRUE),
            sd_pop = sd(population_size, na.rm = TRUE))

```

##Question 1..1 Which whale population is largest? Which is most variable?

### Blue whales are the largest. Gray whales are the most variable based on our data.

#Section 2

##Q2.1 Read in the data

```{r}
US_energy.wide <- read_csv("data/US_energy.wide.csv")
```

##Q2.2a What are the column names of our dataframe?

```{r}
names(US_energy.wide)
```

#### Q2.2b) What unique values are located in the "YYYYMM" column?

```{r}
unique(US_energy.wide$YYYYMM)
```

#### Q2.2c) Are all the units in the "Unit" column the same?

```{r}
#No there are years and months lets separate

library(janitor)

US_energy.wide %>% 
  clean_names() %>% 
  mutate(yyyymm = as.character(yyyymm), .keep = "unused") %>% 
  relocate(yyyymm) %>% 
  separate(col = yyyymm, into = c("yyyy", "mm"), 4) %>% 
  head()

 
separate(US_energy.wide, col = YYYYMM, 
           into = c("YYYY", "MM"), 
           sep = 4)
```
